% chapter01.tex

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %                                                                           %
 %    PyMS documentation                                                     %
 %    Copyright (C) 2005-2012 Vladimir Likic                                 %
 %                                                                           %
 %    The files in this directory provided under the Creative Commons        %
 %    Attribution-NonCommercial-NoDerivs 2.1 Australia license               %
 %    http://creativecommons.org/licenses/by-nc-nd/2.1/au/                   %
 %    See the file license.txt                                               %
 %                                                                           %
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}

PyMS is software for processing of chromatography--mass spectrometry data.
PyMS is written in Python programming language \cite{python}, and is
released as open source, under the GNU Public License version 2. The
driving idea behind PyMS is to provide a framework and a set of
components for rapid development and testing of methods for processing
of chromatography--mass spectrometry data. PyMS provides interactive
processing capabilities through any of the various interactive Python
front ends ("shells"). PyMS is essentially a Python library of
chromatography--mass spectrometry processing functions, and individual
commands can be collected into scripts which then can be run
non-interactively when it is preferable to run data processing in
the batch mode.

PyMS functionality consists of modules which are loaded when needed. 
For example, one such module provides display capabilities, and can
be used to display time dependent data (e.g. total ion chromatogram),
mass spectra, and signal peaks. This module is loaded only when
visualisation is needed. If one is interested only in noise smoothing,
only noise filtering functions are loaded into the Python environment
and used to smooth the data, while the visualisation module (and other
module) need not be loaded at all.

This modularity is supported on all levels of PyMS. For example, if
one is interested in noise filtering with the Savitsky-Golay filter,
only sub-module for Savitsky-Golay filter need to be loaded from
the noise smoothing module, disregarding other modules, as well
as other noise smoothing sub-modules. This organisation consisting
of hierarchical modules ensures that the processing pipeline is
put together from well defined modules each responsible for specific
functions; and furthermore different functionalities are completely
decoupled from one another, providing that implementing a new
functionality (such as test or prototype of a new algorithm) can
be implemented efficiently and ensuring that this will not break
any existing functionality.

\section{The PyMS project}

The PyMS project consists of three parts, each of which exists as a project
in the Google Code repository that can be downloaded separately. The three
parts are:

\begin{itemize}
  \item pyms -- The PyMS code (http://code.google.com/p/pyms/)
  \item pyms-docs -- The PyMS documentation (http://code.google.com/p/pyms-docs/)
  \item pyms-test -- Examples of how to use PyMS (http://code.google.com/p/pyms-test/)
\end{itemize}

The project 'pyms' contains the source code of the Python package PyMS.
The project 'pyms-docs' contains PyMS style guide (relevant for those
who contribute to the PyMS code) and User Guide (this document). The
project 'pyms-test' contains tests and examples showing how to use
various PyMS features. These examples are explained in detail in 
subsequent chapters of this User Guide.

In addition, the current PyMS API documentation (relevant for those
who are interested in PyMS internals) is available from here:\\
{\tt http://bioinformatics.bio21.unimelb.edu.au/pyms.api/index.html}\\

\section{PyMS installation}

PyMS is written in Python, and extensible and modular object-oriented
scripting language \cite{python}. Python is highly portable, cross-platform 
programming language which works well on all major modern operating
systems (Linux, MacOS X, Windows). PyMS is written in pure Python, and
therefore works on all platforms on which Python has been ported.

The PyMS is essentially a python library (a 'package' in python parlance,
which consists of several 'sub-packages'), and some of its functionality
depends on several Python libraries, such as 'numpy' (the Python library
for numerical computing), or 'matplotlib' (the Python library for plotting).
These also need to work on the operating system of your choice for the
functionality they provide to PyMS to be available. In general, the
libraries that PyMS uses are available for all operating systems. The
exception is 'pycdf' - a python interface to Unidata netCDF library
written by Andre Gosselin of the Institut Maurice-Lamontagne, Canada
(http://pysclint.sourceforge.net/pycdf/). {\em This library works only
under Linux/Unix and therefore PyMS functionality which depends on
it works only under the Unix operating system}.

There are several ways to install PyMS depending your computer
configuration and personal preferences. PyMS has been developed
on Linux, and a detailed installation instructions for Linux are
given below. PyMS should work on all major Linux distributions,
and has been tested extensively on Red Hat Linux.

\subsection{\label{subsec:code-download}Downloading PyMS source code}

PyMS source code resides on publicly accessible Google Code servers,
and can be accessed from the following URL: http://code.google.com/p/pyms/.
Under the section "Source" one can find the instructions for downloading
the source code. The same page provides the link under "This project's
Subversion repository can be viewed in your web browser" which allows
one to browse the source code on the server without actually downloading
it. Regardless of the target operating system, the first step towards
PyMS installation is to download the PyMS source code.

Google Code server maintains the source code with the program `subversion'
(an open-source version control system). To download the source code
one needs to use the subversions client. Several subversion clients are
available, some are open source, some freeware, and some are commercial
(for more information see http://subversion.tigris.org/). The svn client
programs are available for all operating systems. For example, on Linux
we use the svn client program which ships with most Linux systems called
simply 'svn'. The 'svn' exists for all mainstream operating
systems\footnote{For example, on Linux CentOS 4 it ships as a part of
the RPM package `subversion-1.3.2-1.rhel4.i386.rpm'}. A well known 
svn client for Windows is TortoiseSVN (http://tortoisesvn.tigris.org/).
TortoiseSVN provides graphical user interface, and is tightly integrated
with Windows. TortoiseSVN is open source and can be downloaded from the
project web page (http://tortoisesvn.tigris.org/). There are also several
commercial svn clients for Windows.

Subversion has extensive functionality, however only the very basic
functionality is needed to download PyMS source code. For more information
about subversion please consult the book freely available
at http://svnbook.red-bean.com/.

If the computer is connected to the internet, and the subversion client
'svn' is installed. On Linux, the following command will download the
latest PyMS source code:

\begin{verbatim}
$ svn checkout http://pyms.googlecode.com/svn/trunk/ pyms
A    pyms/Peak
A    pyms/Peak/__init__.py
A    pyms/Peak/List
A    pyms/Peak/List/__init__.py
... [ further output deleted ] ...
Checked out revision 71.
\end{verbatim}

\section{\label{sec:external-libraries}External Libraries}

In addition to the source code, for the full PyMS functionality
several external libraries are required.

\subsection{\label{subsec:numpy}Package 'NumPy' (required for all
aspects of PyMS)}

The package NumPy is provides numerical capabilities to Python. This
package is used throughout PyMS (and also required for some external
packages used in PyMS), to its installation is mandatory.

The NumPy web site {\tt http://numpy.scipy.org/} provides the installation
instructions and the link to the source code.

\subsection{\label{subsec:pycdf}Package 'pycdf' (required for reading
ANDI-MS files)}

The pycdf (a python interface to Unidata netCDF library) source and
installation instructions can be downloaded from\\
{\tt http://pysclint.sourceforge.net/pycdf/}. Follow the installation
instructions to install pycdf.

\subsection{\label{subsec:pycluster}Package 'Pycluster' (required for peak
alignment by dynamic programming)}

The peak alignment by dynamic programming is located in the subpackage
pyms.Peak.List.DPA. This subpackage uses the Python package `Pycluster'
as the clustering engine. Pycluster with its installation instructions
can be found here:\\
{\tt http://bonsai.ims.u-tokyo.ac.jp/~mdehoon/software/cluster/index.html}.

\subsection{\label{subsec:scipy-ndmage}Package 'scipy.ndimage' (required
for TopHat baseline corrector)}

If the full SciPy package is installed the 'ndimage' will be available.
However the SciPy contains extensive functionality, and its installation
is somewhat involved. Sometimes it is preferable to install only the
subpackage `ndimage'. This subpackage is provided as the PyMS-dependencies
gzipped file available for download from the PyMS webpage (see below).

\subsection{\label{subsec:matplotlib}Package 'matplotlib' (required 
for plotting)}

The displaying of information such as Ion Chromatograms and detected
peaks requires the package matplotlib. The matplotlib package can be
downloaded from:\\
{\tt http://matplotlib.sourceforge.net/}

\subsection{\label{subsec:mpi4py}Package 'mpi4py' (required 
for parallel processing)}

This package is required for parallel processing with PyMS. It can be
downloaded from:\\
{\tt http://code.google.com/p/mpi4py/}

\section{Notes on PyMS installation on Linux}

It is recommended to compile your own Python installation before installing
PyMS. PyMS installation involves placing the PyMS code directory (pyms/)
into a location visible to the Python interpreter. This can be in the
standard place for 3rd party software (the directory site-packages/), or
alternatively if PyMS code is placed in a non-standard location the
Python interpreter needs to be made aware of it before before it is
possible to import PyMS modules. For more on this please consult the
Python command sys.path.append().

In order for existing test scripts provided with the 'pyms-test' project
to run without any modification, PyMS should be installed in the
{\tt /x/PyMS/} directory. To create this directory enter on a command
line:

\begin{verbatim}
$ mkdir -p /x/PyMS/
\end{verbatim}

\section{PyMS installation tutorial (current)}

Currently PyMS is being developed on Linux with the following set of
packages:

\begin{verbatim}
Python-2.7.2
numpy-1.6.1
netcdf-4.1.3
pycdf-0.6-3b
scipy-ndimage.tar.gz
scipy-misc.tar.gz
Pycluster-1.5.0
\end{verbatim}

We provide these packages bundled together into the archive
'PyMS-Linux-deps-2.0.tar.gz' which can be downloaded from
the Bio21 Institute web server at the University of Melbourne:\\
http://bioinformatics.bio21.unimelb.edu.au/pyms.html\\

In addition to the dependencies bundle, one also needs to dowload the
PyMS source code as explained in the section \ref{subsec:code-download}).

Below is a tutorial showing how to install all the packages on Red Hat
Linux 6.

\begin{enumerate}

\item 'Python' installation:

\begin{verbatim}
$ tar xvf Python-2.7.2.tgz
$ cd Python-2.7.2
$ ./configure
$ make
$ make install
\end{verbatim}

\noindent
This installs python in /usr/local/lib/python2.7. It is recommended
to make sure that python called from the command line is the one
just compiled and installed. Starting python from the command line
should report version 2.7:

\begin{verbatim}
$ python
Python 2.7.2 (default, Feb 15 2012, 14:00:58) 
[GCC 4.4.4 20100726 (Red Hat 4.4.4-13)] on linux2
Type "help", "copyright", "credits" or "license" for more information.
>>> 
\end{verbatim}

\item Package 'NumPy' (required for all aspects of PyMS)

The package NumPy is provides numerical capabilities to Python. This
package is used throughout PyMS (and also required for some external
packages used in PyMS), so its installation is mandatory.

\begin{verbatim}
$ tar xvf numpy-1.6.1.tar.gz
$ cd numpy-1.6.1
$ python setup.py install
\end{verbatim}

\item Package 'pycdf'

In order to install {\tt pycdf}, netcdf must first be installed. We
suggest installing netcdf without the HCDF dependency, as this is not
needed for PyMS.

\begin{verbatim}
$ tar xvf netcdf-4.1.3.tar.gz
$ cd netcdf-4.1.3
$ ./configure --disable-netcdf-4
$ make
# make install
\end{verbatim}

Note that by default, netcdf libraries are installed in '/usr/local/lib'.

Now the {\tt pycdf} library can be installed:

\begin{verbatim}
$ tar xvf pycdf-0.6-3b.tar.gz
$ cd pycdf-0.6-3b
$ python setup.py install
\end{verbatim}

It is useful to run the 'pycdf' package example to verify that 'netcdf'
and 'pycdf' libraries are installed correctly. In the folder
'pycdf-0.6-3b/examples' run:

\begin{verbatim}
]$ python compr.py
Running with pycdf configured for: numpy

step 1, obs_table
[[ 1  3 -4  5 10]
 [ 0 10  8  7  4]
 [-1  4  5  9 13]]
should be
[[ 1  3 -4  5 10]
 [ 0 10  8  7  4]
 [-1  4  5  9 13]]

...more output omitted...
\end{verbatim}

If the pycdf example produces an error similar to this:

\begin{verbatim}
$ python compr.py 
Traceback (most recent call last):
  File "copr.py", line 7, in <module>
.....
ImportError: libnetcdf.so.7: cannot open shared object file: No such file
or directory
\end{verbatim}

This indicates tha pycdf cannot find the netcdf shared libraries. 
For example, the netcdf library is installed in '/usr/local/lib',
but may not be cashed and ready for use. This is normally achieved
with the command 'ldconfig'. If the 'netcdf' library is properly
installed and ready to use, the output of 'ldconfig' command
should be similar to the following:

\begin{verbatim}
]# /sbin/ldconfig -v | grep netcdf
        libnetcdf.so.7 -> libnetcdf.so.7.1.1
        libnetcdff.so.5 -> libnetcdff.so.5.1.0
        libnetcdf_c++.so.4 -> libnetcdf_c++.so.4.1.0
\end{verbatim}

If this command produces no output, consult 'ldconfig' configuration
options. In particular, the directory where 'netcdf' was installed
('/usr/local/lib' by default) must be included in the library path.
This can be achieved either by modifying '/etc/ld.so.conf.d' or by
using the {\tt LD\_LIBRARY\_PATH} environment variable. 

\item Package 'PyCluster'

PyCluster is used by PyMS for peak alignment by dynamic programming.

\begin{verbatim}
$ tar xvf Pycluster-1.50.tar.gz
$ cd Pycluster-1.50
$ python setup.py install
\end{verbatim}

\item Package 'SciPY'

The module 'ndimage', used for baseline correction in PyMS, is part
of a much larger library called SciPy. To save installation of such
a large library we install only the parts we require. These are the
ndimage library, and a dependency of ndimage; scipy's 'misc' library.

Modules which require 'ndimage' expect it to be part of this scipy
library so some extra steps are necessary to allow ndimage to be
accessed properly by Python.

First:

\begin{verbatim}
$ tar xvf scipy-ndimage.tar.gz
$ cd ndimage
$ python setup.py install
\end{verbatim}

\begin{verbatim}
$ tar xvf scipy-misc.tar.gz
$ cd misc
$ python setup.py install
\end{verbatim}

The ndimage and misc libraries should be now installed in the
location: {\tt /usr/local/lib/python2.7/site-packages/}.  (If your
version of Python is 2.7). Observe the screen output from the {\tt
  python setup.py install} command to see precisely where they have
been installed.

Python programs will expect ndimage to be at {\tt
  /usr/local/lib/python2.7/site-packages/scipy/ndimage} so we will
manually put it there, and add an \_\_init\_\_.py file which tells
Python that a library is present at that location. We must also move
module 'misc' so that 'ndimage' finds it:

\begin{verbatim}
$ cd /usr/local/lib/python2.7/site-packages
$ mkdir scipy
$ touch scipy/__init__.py
$ mv ndimage scipy
$ mv misc scipy
\end{verbatim}

\end{enumerate}

\subsection{Extra Dependencies}

In addition to the dependencies listed above, PyMS also makes use of
some other python packages for specific needs. The first, matplotlib,
is used to display graphical output (such as ion chromatograms, peak
locations). mpich and mpi4py, together provide the ability to run PyMS
scripts in parallel.

\begin{enumerate}

\item Package 'matplotlib'

\begin{verbatim}
$ tar xvf matplotlib-0.99.1.2.tar.gz
$ cd matplotlib-0.99.1.1  
$ python setup.py build   
$ python setup.py install
\end{verbatim}

\item Installation of the package 'mpi4py'

Before 'mpi4py' can be installed, MPI libraries for parallel computation
(mpich) must be installed.

From the mpich2 project web site download the current distribution of mpich2
(in our case the file 'mpich2-1.2.1p1.tar.gz'), or use the version included
with PyMS-deps versions 2.0 and above.

Prepare the directory for mpich2 installation. In this example we have chosen
to use /usr/local/mpich2/. Our version of mpitch2 is 1.2.1, and to allow for
the installation of different version later, we create a subdirectory "1.2.1",

\begin{verbatim}
$ mkdir -vp /usr/local/mpich2/1.2.1
\end{verbatim}

The above command will make the directory /usr/local/mpich2/ and also
/usr/local/mpich2/1.2.1. Note that /usr/local is usually owned by root,
and the above commands may require root privileges.

Unpack the mpitch source archive and change to the source code directory:

\begin{verbatim}
$ tar xvf mpich2-1.2.1p1.tar.gz 
$ cd mpich2-1.2.1p1
\end{verbatim}

Configure, compile, and install mpich2:

\begin{verbatim}
$ ./configure --prefix=/usr/local/mpich2/1.2.1 --enable-sharedlibs=gcc
$ make
$ make install
\end{verbatim}

If /usr/local/mpich2/1.2.1 is owned by root, the last command may require
root privileges.

With this completed, one can install the package 'mpi4py'. From the mpi4py
project web site download the current distribution of mpi4py (in our case
the file 'mpi4py-1.2.1.tar.gz'), or use the distribution included in
PyMS-Deps 2.0 and above.  Unpack the source archive and change to the
source code directory:

\begin{verbatim}
$ tar xvf mpi4py-1.2.1.tar.gz
$ cd mpi4py-1.2.1
\end{verbatim}

Edit the file 'mpi.cfg' which can be found in this directory, to reflect the
location of mpich2.  In our case this file after editing contained the
following:

\begin{verbatim}
# MPICH2
[mpi]
mpi_dir              = /usr/local/mpich2/1.2.1
mpicc                = %(mpi_dir)s/bin/mpicc
mpicxx               = %(mpi_dir)s/bin/mpicxx
\end{verbatim}

Install the package 'mpi4py':

\begin{verbatim}
$ python setup.py install
\end{verbatim}

Check that mpi4py works:

\begin{verbatim}
]$ python
Python 2.7.2 (default, Feb 15 2012, 14:00:58) 
[GCC 4.4.4 20100726 (Red Hat 4.4.4-13)] on linux2
Type "help", "copyright", "credits" or "license" for more information.
>>> import mpi4py
>>> 
\end{verbatim}

If the above command import produced no output, mpi4py is installed
properly and ready to use.

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Old dependencies installation tutorial
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{PyMS installation tutorial (old)}

Prior to 2011, PyMS was being developed on Linux with the following set of
packages:

\begin{verbatim}
Python-2.5.2
numpy-1.1.1
netcdf-4.0
pycdf-0.6-3b
ndimage.zip
Pycluster-1.41
matplotlib-0.99.1.2
mpi4py-1.2.1.tar.gz
mpich2-1.2.1p1.tar.gz
\end{verbatim}

We provide these packages bundled together into the archive
'PyMS-Linux-deps-1.1.tar.gz' which can be downloaded from
the Bio21 Institute web server at the University of Melbourne:\\
http://bioinformatics.bio21.unimelb.edu.au/pyms.html\\

In addition to the dependencies bundle, one also needs to dowload the
PyMS source code as explained in the section \ref{subsec:code-download}).

Below is a tutorial showing how to install all the packages on Red Hat
Linux 5.

\begin{enumerate}

\item 'Python' installation:

\begin{verbatim}
$ tar xvfz Python-2.5.2.tgz
$ cd Python-2.5.2
$ ./configure
$ make
$ make install
\end{verbatim}

\noindent
This installs python in /usr/local/lib/python2.5. It is recommended
to make sure that python called from the command line is the one
just compiled and installed.

\item 'NumPy' installation:

\begin{verbatim}
$ tar xvfz numpy-1.1.1.tar.gz
$ cd numpy-1.1.1
$ python setup.py install
\end{verbatim}

\item 'pycdf' installation

Pycdf has two dependencies: the Unidata netcdf library and NumPy. The
NumPy installation is described above. To install pycdf, the netcdf
library must be downloaded\\
({\tt http://www.unidata.ucar.edu/software/netcdf/index.html}),\\
compiled and installed first:

\begin{verbatim}
$ tar xvfz netcdf.tar.gz
$ cd netcdf-4.0
$ ./configure
$ make
$ make install
\end{verbatim}

The last step will create several binary `libnetcdf*' files in
/usr/local/lib. Then 'pycdf' should be installed as follows:

\begin{verbatim}
$ tar xvfz pycdf-0.6-3b
$ cd pycdf-0.6-3b
$ python setup.py install
\end{verbatim}

\item 'Pycluster' installation

\begin{verbatim}
$ tar xvfz Pycluster-1.42.tar.gz
$ cd Pycluster-1.42
$ python setup.py install
\end{verbatim}

\item 'ndimage' installation:

\begin{verbatim}
$ unzip ndimage.zip
$ cd ndimage
$ python setup.py install --prefix=/usr/local
\end{verbatim}

\noindent
Since 'ndimage' was installed outside the scipy package, this requires
some manual tweaking:

\begin{verbatim}
$ cd /usr/local/lib/python2.5/site-packages
$ mkdir scipy
$ touch scipy/__init__.py
$ mv ndimage scipy
\end{verbatim}

\item 'matplotlib' installation:

\begin{verbatim}
$ tar xvfz matplotlib-0.99.1.2
$ cd matplotlib-0.99.1.1
$ python setup.py build
$ python setup.py install
\end{verbatim}

\noindent
The 'pyms.Display' module uses the TKAgg backend for matplotlib. If
this is not your default backend, the matplotlibrc file may be edited.
To locate the file 'matplotlibrc', in a python interactive session:

\begin{verbatim}
>>> import matplotlib
>>> matplotlib.matplotlib_fname()
\end{verbatim}

\noindent
Open the matplotlibrc file in a text editor and adjust the 'backend'
parameter to 'TKAgg'.

\item 'mpi4py' installation:

This package is required for running PyMS processing on multiple processors
(CPUs). Instructions how to install this package and run PyMS processing
in parallel are given in Section \ref{sec:mpi}. 

\end{enumerate}

\section{Troubleshooting}

The most likely problem with PyMS installation is a problem with
installing one of the PyMS dependencies.

\subsection{Pycdf import error}

On Red Hat Linux 5 the SELinux is enabled by default, and this causes the
following error while trying to import properly installed pycdf:

\begin{verbatim}
$ python
Python 2.5.2 (r252:60911, Nov  5 2008, 16:25:39)
[GCC 4.1.1 20070105 (Red Hat 4.1.1-52)] on linux2
Type "help", "copyright", "credits" or "license" for more information.
>>> import pycdf
Traceback (most recent call last):
  File "<stdin>", line 1, in <module>
  File "/usr/local/lib/python2.5/site-packages/pycdf/__init__.py",
        line 22, in <module> from pycdf import *
  File "/usr/local/lib/python2.5/site-packages/pycdf/pycdf.py",
        line 1096, in <module> import pycdfext as _C
  File "/usr/local/lib/python2.5/site-packages/pycdf/pycdfext.py",
        line 5, in <module> import _pycdfext
ImportError: /usr/local/lib/python2.5/site-packages/pycdf/_pycdfext.so:
    cannot restore segment prot after reloc: Permission denied
\end{verbatim}

This problem is remedied by disabling SELinux (login as `root', open the
menu Administration $\rightarrow$ Security Level and Firewall, tab SELinux,
change settings from `Enforcing' to `Disabled').

This problem is likely to occur on Red Hat Linux derivative distributions
such as CentOS.

\section{PyMS tutorial and examples}

This document provides extensive tutorial on the use of PyMS, and the
corresponding examples can be downloaded from the publicly accessible
Google code project 'pyms-test' (http://code.google.com/p/pyms-test/). 
The data used in PyMS documentation and examples is available from the
Bio21 Institute server at the University of Melbourne. Please follow
the link from the PyMS web page:\\
{\tt http://bioinformatics.bio21.unimelb.edu.au/pyms}\\
or go directly to\\
{\tt http://bioinformatics.bio21.unimelb.edu.au/pyms-data/}\\

A tutorial illustrating various PyMS features in detail is provided
in subsequent chapters of this User Guide. The commands executed
interactively are grouped together by example, and provided as
Python scripts in the project 'pyms-test' (this is a Google code
project, similar to the project 'pyms' which contains the PyMS
source code).

The setup used in the examples below is as follows. The projects 'pyms',
'pyms-test', 'pyms-docs', and 'data' are all in the same directory,
'/x/PyMS'. In the project 'pyms-test' there is a directory corresponding
to each example, coded with the chapter number (ie. {\tt pyms-test/21a/}
corresponds to the Example 21a, from Chapter 2).

In each example directory, there is a script named 'proc.py' which
contains the commands given in the example.  Provided that the paths
to 'pyms' and 'pyms-data' are set properly, these scripts could be
run with the following command:

\begin{verbatim}
$ python proc.py
\end{verbatim}

Before running each example the Python interpreter was made aware of
the PyMS location with the following commands:

\begin{verbatim}
import sys
sys.path.append("/x/PyMS")
\end{verbatim}

For brevity these commands will not be shown in the examples below,
but they are included in 'pyms-test' example scripts.  The above
path may need to be adjusted to match your own directory structure.

\section{\label{sec:pyms-windows}Using PyMS on Windows}

Python is highly cross-platform compatible, and PyMS works seamlessly
on Windows. The only exception is reading of data in ANDI-MS (NetCDF)
format, widely used standard format for storing raw
chromatography--mass spectrometry data (see \ref{sec:chapter2-introduction})
This capability in PyMS depends on the 'pycdf' library, which is not
supported on Windows (see the Subsection \ref{subsec:pycdf}). Therefore
at present the ability to read ANDI-MS files is limited to Linux. All
other PyMS functionality is available under Windows.

We use Linux for the development and deployment PyMS, and this User
Guide largely assumes that PyMS is used under Linux. In this section
we give some pointers on how to use PyMS under Windows.

\subsection{PyMS installation on Windows}

\begin{enumerate}

\item Install Python, NumPy, SciPy, and matplotlib for Windows.
The bundle of these software packages tested under Windows XP and
Windows 7 can be downloaded from the PyMS project page at the
Bio21 Institute, University of Melbourne\\
{\tt http://bioinformatics.bio21.unimelb.edu.au/pyms.html}

\item Download the latest PyMS code from the PyMS Google Code
project page\\
{\tt http://code.google.com/p/pyms/}

\item Unpack the PyMS code and place it in a standard place for
Python libraries, or adjust the PYTHONPATH variable to include
the path to PyMS. If Python is installed in C:\\Python25, the
standard place for Python libraries is C:\\Python25\\Libs\\site-packages

\item Start IDLE, or other Python shell. If PyMS is installed
properly, the following command will not return any warning or
error messages:

\begin{verbatim}
>>> import pyms
\end{verbatim}

\end{enumerate}

\subsection{Example processing GC-MS data on Windows}

The example data can be downloaded from\\
{\tt http://bioinformatics.bio21.unimelb.edu.au/pyms/data/}.
We will use the data file in the JCAMP-DX format, named
{\tt gc01\_0812\_066.jdx}. Once downloaded this data file
will be placed in the folder {\tt C:Data} for the example
below.

The Python environment can be accessed from several Python
shells. The default shell that comes with the Python 2.5
installation is IDLE. In this example we first load the
raw data,

\begin{verbatim}
>>> from pyms.GCMS.IO.JCAMP.Function import JCAMP_reader
>>> jcamp_file = "C:\Data\gc01_0812_066.jdx"
>>> data = JCAMP_reader(jcamp_file)
 -> Reading JCAMP file 'C:\Data\gc01_0812_066.jdx'
\end{verbatim}

The intensity matrix object is built by binning:

\begin{verbatim}
>>> from pyms.GCMS.Function import build_intensity_matrix_i
>>> im = build_intensity_matrix_i(data)
\end{verbatim}

In this example, we show how to btain the dimensions of the
newly created intensity matrix, then loop over all ion chromatograms,
and for each ion chromatogram apply Savitzky-Golay noise filter
and tophat baseline correction:

\begin{verbatim}
>>> n_scan, n_mz = im.get_size()
>>> from pyms.Noise.SavitzkyGolay import savitzky_golay
>>> from pyms.Baseline.TopHat import tophat
>>> for ii in range(n_mz):
    print "working on IC", ii
    ic = im.get_ic_at_index(ii)
    ic1 = savitzky_golay(ic)
    ic_smooth = savitzky_golay(ic1)
    ic_base = tophat(ic_smooth, struct="1.5m")
    im.set_ic_at_index(ii, ic_base)
\end{verbatim}

The resulting noise and baseline corrected ion chromatogram is saved
back into the intensity matrix. As this example illustrates, PyMS
depends on Python and can be used in the exactly the same way under
any operating system that supports Python. For more advanced use
refer to subsequent chapters.

